---
output: github_document
bibliography: Appendix_bib.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
<!-- badges: end -->

# HerdDynamics
The goal of HerdDynamics is to provide a set of scripts to simulate the growth of livestock herds under various management strategies.

## Installation

You can install the development version of HerdDynamics from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("penguinnick/HerdDynamics")
```

## Introduction - Livestock Population Dynamics and Stochastic Dynamic Programming

This document outlines the procedures for simulating goat and sheep herd dynamics and using the resulting models in a stochastic dynamic program that will identify the optimal culling strategy and herd reproduction parameters. The code here supports Chapter 2 of Nick Triozzi's PhD Dissertation.

This is a basic example which shows you how to project herd growth under a single offtake strategy:

```{r example, echo = FALSE}
library(HerdDynamics)
library(tidyverse)
```

## Age Classes

The first step is to build a table containing information about the age groups used in the projections. In this program age classes for males and females may be specified separately. However, culling profiles constructed from archaeological remains cannot distinguish male from female mandibles. Therefore, offtake rates modeled here will be applied to the entire herd while intrinsic mortality rates will be defined separately for males and females.

```{r ageclass}

# The age classes are defined based on Payne's (1973) age classes for sheep and goats.
HerdDynamics::Payne_ages

#-- create the tcla table, containing age classes for each sex, and the starting and ending ages for each class in months
tcla <- build_tcla( female.ages = Payne_ages$lclass/12, 
                    male.ages = Payne_ages$lclass/12, 
                    nbphase = 12 #The parameter nbphase is set to 12 which converts ages to months.
                    ) 
tcla
```
# Offtake

A set of offtake rates are included in the package as `offtake_models`. This list contains survivorship rates for standardized  age classes following Marom and Bar-Oz [-@Marom2009].  These offtake rates usually describe culling strategies for a whole herd, regardless of sex. Here we use them to model the offtake rates of females. For now, we will focus only on Payne's [-@Payne1973] meat, milk, and wool strategies.

```{r offtake}

#-- subset offtake models list for Milk, Meat, and Wool
offtake.models = HerdDynamics::offtake_models[c("Milk", "Meat", "Wool")]

#-- convert survivorship to mortality
offtake.models <- lapply(offtake.models, function(x) { 1 - (x / 100) })

#-- plot offtake models
plot_offtake(offtake.models, title = "Culling Strategies for Sheep and Goats")
```

## Herd Population Growth Parameters - Intrinsic Mortality

The probability that an animal will survive from one timestep to the next is affected by the competing risks of being slaughtered (i.e., offtake) and intrinsic mortality. We set intrinsic mortality for males and females separately.

```{r mortality}
#-- goat intrinsic mortality rates
doe.mortality =  c( 0.179, 0.453, 0.18, 0.18, 0.18, 0.18,  0.18,  0.18, 0.18)
buck.mortality = c(0.179,  0.453, 0.15, 0.15, 0.05, 0.05,  0.05,  1.00, 1.00)
```

## Fertility and Mortality Parameters
Several parameters are important regarding the reproductive biology of goats and sheep.
*part.age* specifies the age of first parturition. 
*parturition* specifies the number of parturitions per female per year. 
*prolificacy* specifies the prolificacy rate, defined as the number of live offspring per parturition per year. Age-specific rates are available.


```{r calculate-transition-matrix}
#-- goat reproduction and mortality parameters
goat.parms = list(
  ages= c( Payne_ages$ages),
  parturition = 1.2,
  part.age = 2,  # age of first parturition
  prolificacy = c( 0, 0, 0, 0.82, 1.10, 1.41, 1.45, 1.03, 1.03), # number of livebirths expected per animal per parturition per year by age class
  f.mortality = doe.mortality,
  m.mortality = buck.mortality
  # f.offtake = doe.mortality/2
)

#-- example of how to set the parameters for a single offtake strategy
build_param(tcla = tcla, 
            parms = goat.parms, 
            Inf.Mortality = "auto", # setting this variable to "auto" will use the offtake model to calculate infant mortality rates based on the mean and standard deviation of mortality rates for animals under 1 year of age.
            prolificacyRate = "auto", # setting this variable to "auto" will calculate age-specific prolificacy rates based on the mean and standard deviation of prolificacy rates provided in the goat.parms list.
            offtake = offtake.models$Milk, 
            nbphase = 12 )

#-- create a list of parameters for each offtake strategy
param = lapply(offtake.models, function(o) { 
  build_param(tcla = tcla, 
              parms = goat.parms, 
              Inf.Mortality = "auto", 
              prolificacyRate = "auto", 
              nbphase = 12,
              offtake = o)
  })

#-- inspect Milk model transition matrix parameters
param$Milk$param
```

## Reproductive Values
After setting these variables we create a parameter data frame that will be used to compute lambda, reproductive values, and project herd growth. The function *get_lambda* will calculate the reproductive values for each age class and the overall lambda value (growth rate) for the herd.

```{r reproduction}
#-- get reproductive parameters using the get_lambda function with an initial herd size of 1000 animals
milk.lambda = get_lambda(param$Milk$param, tcla = tcla, p0=1000)

#-- herd growth rate
milk.lambda$lambda

#-- sex proportion of herd
milk.lambda$sex.proportion

#-- initial age structure of herd
milk.lambda$initial.herd %>% 
  mutate(ages = rep(Payne_ages$ageClasses, 2)) %>%
  ggplot(aes(x = ages, y = xini, group = sex)) +
  geom_col(aes(fill = sex)) + 
  coord_flip() +
  theme_minimal()

#-- get reproductive values for all offtake strategies
repro = lapply(param, function(p){ get_lambda(p$param, tcla, p0=1000) }) 

```

## Demographic Projection
Now we run the projection. We start with an initial herd size of 200 animals.
First we define *nbcycle* as number of cycles to project through (i.e., years). We keep the previously defined value  *nbphase* (nbphase=12=12 months). *nbstep* is the product of these values and will calculate the change in herd demography from one phase to the next (i.e., from 1 month to the next).

lapply is used to run the function *project_herd* and stores the projection *results* as a list.

```{r projection}
#-- put all parameters into a list
all.param = param_list(param, repro)
str(all.param$Milk, 2)

#-- set the number of cycles and phases
nbcycle = 20 # 20 years
nbphase = 12 # in monthly timesteps

#-- project herd growth
results = lapply(all.param, function(a){ project_herd(a, 20, 12)})

```

```{r plot, echo = FALSE}
#-- summarize results by cycle, time, and sex 
tot.pop.res = lapply(results, function(r){pop_summary(r, tim, cycle, sex)})
z = res2df(tot.pop.res)

p1 = ggplot(data = z, ggplot2::aes(x = tim, y = n)) +
  geom_col(ggplot2::aes( fill=sex), position = "stack") +
  labs(x = "Time",y = "Nb. animals") + 
  facet_wrap(~strategy, nrow = 3) + 
  theme_minimal()
p1

#-- plot population change through time
tot.pop.res = lapply(results, function(r){pop_summary(r, tim)})
z = res2df(tot.pop.res)
p2 = ggplot(data = z, aes(x = tim, y = n, group = strategy)) +
  geom_line(ggplot2::aes(colour = strategy)) +
  labs(x = "Time",y = "Nb. animals") + 
  theme_minimal()
p2
```


# References

